Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2024-08-14 paper/paper.qmd
mild edits

Students from a range of years took the course, however the majority of respondents were in their 3rd or 4th year of study (@tbl-year-dist). The course's prerequisite two-course sequence is typically completed by students in their second year, would make it difficult to take this course earlier than their 3rd year.

@@

writing_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

@@

coding_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")
```

@@

To understand the role of LLMs in learning, students identified their usage in a more granular way by selecting various pre-defined use cases in the survey (@fig-AI-use-and-usefulness-1). Technical questions and explaining concepts were the two top use cases among students in the course. More than half of students also used LLMs for quick questions, general knowledge, writing paper code, and checking solutions. Just less than half used it to write paper content, which could suggest that students do not feel confident using it to improve writing.

@@

The goal of our modelling strategy is to better understand how a respondents' result on their final paper associates with their self-reported LLM usage. Students received their result on the final paper as a proportion (a 0-100% mark), so we use zero-one-inflated beta regression to best fit the distribution in this interval. We use this type of regression given the ability of the beta distribution to effectively describe continuous and skewed or heteroskedastic datasets, which we can see in the asymmetrical histogram of the distribution of marks.

Transforming the regression with a logit function as the link function ensures that the predicted values stay between zero and one, respecting the boundaries of the data. Some students got full marks for the paper, so we use the zero-one-inflated beta model.

Here we briefly describe the model that we use, which follows @kurztotherescue. Model diagnostics are included in [Appendix -@sec-model-details].

@@

LLM usage estimates are in relation to "Extensive" LLM usage. In the base model, the coefficients suggest that less usage of LLMs was associated with lower scores. The second model retains this slight association, but with slightly smaller coefficients. For instance, the negative impact of “b_llm_usageSomewhat” is slightly reduced from -0.35 to -0.29, indicating that part of the effect observed in the base model might be explained by the GPA variable.

>>>>

Students from a range of years took the course, however the majority of respondents were in their 3rd or 4th year of study (@tbl-year-dist). The course's prerequisite two-course sequence is typically completed by students in their second year, which makes it difficult for students to take this course earlier than their 3rd year.

@@

# so that "Very different from me" is Red - better visuals
set1_palette <- c("#4DAF4A", "#377EB8", "#E41A1C")

writing_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_manual(values = set1_palette) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

@@

coding_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_manual(values = set1_palette) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")
```

@@

To understand the role of LLMs in learning, students identified their usage in a more granular way by selecting various pre-defined use cases in the survey (@fig-AI-use-and-usefulness-1). Technical questions and explaining concepts were the two top use cases among students in the course. More than half of students also used LLMs for quick questions, general knowledge, writing paper code, and checking solutions. Just less than half used it to write paper content, which could suggest that students do not feel confident using it to improve writing; this could be similar to the attitudes expressed by students surveyed in @Cahill_2024.

@@

The goal of our modelling strategy is to better understand how a respondents' result on their final paper associates with their self-reported LLM usage. Students received their result on the final paper as a proportion (a 0-100% mark), so we use a Bayesian zero-one-inflated beta regression to best fit the distribution in this interval. We use this type of regression given the ability of the beta distribution to effectively describe continuous and skewed  (heteroskedastic) datasets, which we can see in the asymmetrical histogram of the distribution of marks.

Transforming the regression with a logit function as the link function ensures that the predicted values stay between zero and one, respecting the boundaries of the data.

We briefly describe the model that we use, which follows @kurztotherescue. No students got zeroes, though some got full marks (100%) one the paper, so we adapt the regression to one-inflated beta model as described by @heisstotherescue to account for this.

Model diagnostics are included in [Appendix -@sec-model-details].

@@

LLM usage estimates are in relation to "Extensive" LLM usage. In the base model, the coefficients suggest that less usage of LLMs was associated with lower scores. The second model retains this association, but with slightly smaller coefficients. For instance, the negative impact of “b_llm_usageSomewhat” is slightly reduced from -0.35 to -0.29, indicating that part of the effect observed in the base model might be explained by the GPA variable.
